Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex

\begin{abstract}

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) instead of a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages Conditional Independence
	(CI) testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	Empirical results show that this iterative approach achieves
	performance comparable to automated algorithms if the orientation of
	new edges is correctly determined in at least one out of two cases.
	Additionally, we show that, in the absence of an expert, a Large
	Language Model (LLM) can be used to determine edge orientations. We
	provide an implementation of our approach as both a web tool: <redacted
	for review> and a Python package <redacted for review>.

\end{abstract}

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001} and Fast
Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Researchers often prefer
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We attribute this preference to several challenges with existing
causal discovery algorithms in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Their output can vary significantly depending on
		the choice of algorithm and hyperparameters, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. 
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. 
\end{enumerate}

Figure~\ref{fig:intro} highlights some of these issues. On the other hand,
constructing DAG based solely on domain expertise can be error prone as it
requires us to distinguish between different causal structures, such as direct
or indirect effects or causal mediation. Given these complexities, it is
important to validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the CI statements implied by the
DAG hold in the data \citep{Ankan2021}. Specifically, each missing edge between
a pair of variables in the DAG implies a CI statement, which can be tested in
the dataset using a statistical test. If we find any violation to CI
statements, we can make modifications to our model accordingly to fix the
violations.

>>>>

\begin{abstract}
	Numerous causal discovery algorithms were developed to automatically learn
	directed acyclic graphs (DAGs) and other causal models from data. However,
	their adoption in applied domains remains limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to produce results that contradict
	obvious domain knowledge and their inability to distinguish Markov equivalent
	models. To assist researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages conditional independence
	testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	In a simulation study, we find that this iterative approach to leverage domain 
	knowledge already starts outperforming purely data-driven structure learning if 
	the orientation of new edge is correctly determined in at least one out of two cases.
	We present a proof-of-concept implementation using a large language 
	model as a domain expert and a graphical user interface designed to 
	assist human experts with DAG construction.
\end{abstract}

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001,KalischB07} 
and Fast Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NOTEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Challenges encountered with existing
causal discovery algorithms in practice include but are not limited to:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While constraint-based algorithms are 
		often asymptotically
		consistent \citep{KalischB07}, they can and do make mistakes
		on  finite samples. These mistakes can be severe and contradict
		obvious domain knowledge (think of edges going into unmodifiable
		attributes such as Age). The choice of algorithm and hyperparameters
		significantly aa=ffect the output, making it
		difficult to assess reliability. Additionally, the absence of
		robust performance evaluation methods for any given dataset
		further reduce the confidence in their outputs. A recent paper
		therefore advised Epidemiologists to not attempt using structure
		learning algorithms without help of an expert \citep{Gururaghavendran_2024}.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, automated 
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. This structural uncertainty
		can make difficult or impossible to apply the learned model for downstream
		tasks, such as identification or causal
		effect estimation \citep{Maathuis_2009,PerkovicTKM17}.
\end{enumerate}

Figure~\ref{fig:intro} highlights some of these issues. In practice, DAGs 
are still largely constructed from domain knowledge alone 
\citep{Tennant2020,Petersen2021}. This can, however, be equally problematic.
Constructing DAGs requires us to distinguish between different causal structures, 
such as direct or indirect effects or causal mediation. Specifically, there is often a lack
of theoretical support for the \emph{absence} of direct causal effects, which are 
the key assumptions that graphical models make to enable downstream causal inferences.
Given the likelihood of making mistakes, it is therefore
important to at least validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the conditional independence 
(CI) statements implied by the DAG hold in the data \citep{Ankan2021}. Specifically, 
each missing edge between
a pair of variables in the DAG leads to one or more CI statements, 
which can be checked using statistical tests. Violations to CI
statements can point to the erroneous omission of important causal effects from a model
or to the presence of latent confounders.
